Abstractions for Multi-Sorted Substitutions

### 2023-09-19 paper/paper.tex
Update.

  Additionally, repetition can also occur within a single formalization.
  This can be seen with polymorphic languages, where multiple sorts
  of variables are present.
  Consider for example System F, which supports both expression- and
  type-variables.
  With a naive approach, the whole substitution machinery needs to be
  duplicated three times! We need to substitute expression-variables in
  expressions, type-variables in types, but also type-variables in
  expressions.
  Even worse, having two substitutions acting on expressions, requires
  to also prove lemmas about their interactions.
  If we would additionally introduce kind-variables, we would end up
  with a total of six duplications of the substition machinery and
  corresponding interaction lemmas!

@@

  % In summary, to have enough substitution machinery to prove subject
  % reduction for a language with expression-, type-, and
  % kind-variables one can end up with a total of 24
  % repetitions for each definition related to substitution, not counting
  % the interaction lemmas between different substitutions that are
  % acting on the same syntactic categories!
  If one is not careful, a formalization of a language with type and
  kind polymorphism can easily end up with 24 slightly changed copies of
  the whole substitution machinery, which for de Bruijn indices tends to
  include at least a type for parallel substitution, and operations for
  application, extension, weakening, lifting, singleton substitution,
  extensional equality, and an identity substitution - and this is not
  counting the slightly changed copies of interaction lemmas!

  Clearly, this situation is in need of automation.
  Our framework approaches this problem using a combination of
  abstractions and reflection in the context of extrinsically typed,
  intrinsically scoped syntax with de Bruijn indices.
  The user can write a language specification using regular Agda
  definitions (no generic encodings) and our framework derives
  definitions and lemmas for untyped substitution, and provides an
  abstraction to prove type preservation for all substitution operations
  using a single lemma. For System F, our framework allows to prove
  subject reduction with only a single handwritten lemma for type
  preservation.

  Compared to many other approaches, we do not derive substitutions
  for each of the variable sorts and syntactic categories, but instead
  use a novel approach for defining syntax, which directly supports
  substitutions that replace all variable sorts in parallel and can be
  applied to all syntactic categories.
  By further unifying renamings and substitutions, we gain the ability
  to talk abstractly about any kind of renaming or substitution that
  can occur in the formalization.
  Having unified the untyped substitution machinery, we build further
  abstractions to similarly unify typing and type preservation lemmas.
  Our reflection algorithm instanciates those abstractions

  Together with abstractions to unify renamings and substitutions,
  this allows us to create an abstraction that is general enough to
  prove type preservation for all sub

@@

  In this paper we present a simplified version of the actual
  framework, focusing on the core concepts. We present all necessary
  definitions and lemmas, but omit some proofs for the sake of
  space. The omitted proofs can be found in Appendix~\ref{sec:proofs}
  and in the supplementary material. The full framework can be found at
  TODO.

  The rest of this paper is structured as followed:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to introduce the notion of types
  for multi-sorted syntax and how to define typing contexts.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservattion lemmas for renamings and substitutions.
  Section~\ref{sec:semantics} introduces a small-step operational semantics for System F.
  Section~\ref{sec:subjectreduction} proves the subject reduction lemma for System F.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} describes case studies of object languages beyond System F.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:future} describes future work.
  Section~\ref{sec:conclusion} concludes.

  \subsection{Contributions}
  \label{sec:introduction:contributions}
  \begin{enumerate}
  \item
    a novel approach for formalizing intrinsically-scoped syntax with
    multiple variable sorts as a special kind of intrinsically-typed
    syntax, we call \emph{multi-sorted syntax};
  \item
    an adaptation of McBride's kit-abstraction for unifying renamings
    and substitutions to multi-sorted syntax;
  \item
    a novel abstraction for composition and it's metatheory, unifying
    the four compositions between renamings and substitutions;
  \item
    a novel abstraction for typing relations, unifying type
    preservation of renaming and substitution;
  \item
    a formalized specification of a large class of object languages
    for which substitution and lemmas can be derived generically.
  \item
    an implementation of the concepts as framework in Agda featuring
    a reflection algorithm, representation independence for
    substitutions and typing contexts, heterogeneous equality between
    renamings and substitutions, and free of postulated axioms.
  \item
    three case studies using our approach to prove subject reduction
    for System~F with subtyping, a dependently-typed lambda calculus, and
    pattern matching.
  \end{enumerate}

@@

  \section{Semantics}
  \label{sec:semantics}
  We define a standard small-step operational semantics for System F
  with full reduction:
  \AppCode\FReduction

  \section{Subject Reduction}
  \label{sec:subjectreduction}
  Proving subject reduction is now straightforward:
  \AppCode\FSubjectReduction
  \AppCode\FSubjectReductionProofInteresting
  We only listed the cases that do not follow directly from the
  induction hypothesis:
  \begin{itemize}
  \item The beta-reduction rule for type-application, \ACode{\ACon{β-Λ}},
    is a direct consequence of substitution preserves typing.
    \ACode{⊢e₂} is the typing for \ACode{e₂}.
    \ACode{⊢⦅ ⊢e₂ ⦆} is the substitution-typing for the singleton substitution 
    \ACode{\AField{⦅ e₂ ⦆}}.
    \ACode{\AField{⊢e₁ ⊢⋯ ⊢⦅ ⊢e₂ ⦆}} states that applying the singleton substitution
    to \ACode{e₁} preserves typing.
  \item The beta-reduction rule for value-application is almost the same,
    but requires us to addionaly apply an equality to the type index.
  \end{itemize}

>>>>

  Additionally, repetition can also occur within a single formalization.
  This can be seen with polymorphic languages, where multiple sorts
  of variables are present.
  Consider for example System F, which supports both expression- and
  type-variables.
  With a naive approach, the whole substitution machinery needs to be
  duplicated three times! We need to substitute expression-variables in
  expressions, type-variables in types, but also type-variables in
  expressions.
  Even worse, having two substitutions acting on expressions requires
  to also prove lemmas about their interactions.
  If we would additionally introduce kind-variables, we would end up
  with a total of six duplications of the substition machinery and
  corresponding interaction lemmas!

@@

  % In summary, to have enough substitution machinery to prove subject
  % reduction for a language with expression-, type-, and
  % kind-variables one can end up with a total of 24
  % repetitions for each definition related to substitution, not counting
  % the interaction lemmas between different substitutions that are
  % acting on the same syntactic categories!
  If one is not careful, a formalization of a language with type and
  kind polymorphism can easily end up with 24 slightly changed copies of
  the whole substitution machinery, which for de Bruijn indices tends to
  include at least a type for parallel substitution, and operations for
  application, extension, weakening, lifting, singleton substitution,
  extensional equality, and an identity substitution - and this is not
  counting the slightly changed copies of interaction lemmas!
  Clearly, this situation is in need of automation.

  Our framework approaches this problem using a combination of
  abstractions and reflection in the context of extrinsically typed,
  intrinsically scoped syntax with de Bruijn indices.
  The user can write a language specification using regular Agda
  definitions (no generic encodings) and our framework derives
  definitions and lemmas for untyped substitution, and provides an
  abstraction to prove type preservation for all substitution operations
  using a single lemma. For System F, our framework allows to prove
  subject reduction with only a single handwritten lemma for type
  preservation.

  Compared to many other approaches, we do not derive substitutions
  for each of the variable sorts and syntactic categories, but instead
  use a novel approach for defining syntax, which directly supports
  substitutions that replace all variable sorts in parallel and can be
  applied to all syntactic categories.
  By further unifying renamings and substitutions, we gain the ability
  to talk abstractly about any kind of renaming or substitution that
  can occur in the formalization.
  This generality is key to then define abstractions for typing and
  type preservation on the same level of generality, allowing to prove
  type preservation for renamings and substitutions of all variable
  sorts and syntactic categories in a single lemma for many typing
  relations, including those of our case studies.

  % TODO: avoiding interaction lemmas

  % By using extrinsic typing, we clearly separate the definitions and
  % lemmas of untyped substitutions from typing and type preservation.
  % This is useful, because the former can be derived for virtually any
  % object language, while the latter cannot, as typing relations can
  % be arbitrarily complex.

  % This does not only separate specification from proofs, making it
  % easier for a human to verify if a specification is correct without
  % understanding the proofs, but also separates things which can be
  % derived for virtually any object language, from things which cannot.

  % This allows untyped substitution to be derived for virtually all
  % object languages, while still allowing for general abstractions
  % for type preservation to be built on top, which cannot be proved
  % automatically in general, as typing relations can become arbitrarily
  % complex.

  % The class of supported typing relations is currently limited to 
  % a subclass of the classical tertiary relations between typing
  % context, term, and type, but strong enough work for our three case studies
  % of System F with subtyping, lambda calculus with dependent function
  % types, and pattern matching.

@@

  In this paper we present a simplified version of the actual
  framework, focusing on the core concepts. We present all necessary
  definitions and lemmas, but omit some proofs for the sake of
  space. The omitted proofs can be found in Appendix~\ref{sec:proofs}
  and in the supplementary material.
  The full framework is available online.\footnote{See supplementary material. In the final version here will be a link to github.}

  The rest of this paper is structured as followed:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to introduce the notion of types
  for multi-sorted syntax and how to define typing contexts.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservattion lemmas for renamings and substitutions.
  Section~\ref{sec:subjectreduction} introduces a small-step
  operational semantics for System F and proves subject reduction.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} discusses our case studies.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:future} discusses future work.
  Section~\ref{sec:conclusion} concludes.

  \subsection{Contributions}
  \label{sec:introduction:contributions}
  \begin{enumerate}
  \item
    a novel approach for formalizing intrinsically-scoped syntax with
    multiple variable sorts as a special kind of intrinsically-typed
    syntax, we call \emph{multi-sorted syntax};
  \item
    an adaptation of McBride's
    kit-abstraction\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
    for unifying renamings and substitutions to multi-sorted syntax;
  \item
    a novel abstraction for composition and it's metatheory, unifying
    the four compositions between renamings and substitutions;
  \item
    a novel abstraction for typing relations, unifying type
    preservation of renaming and substitution;
  \item
    a formalized specification of a large class of object languages
    for which substitution and lemmas can be derived generically.
  \item
    an implementation as an Agda framework featuring
    a reflection algorithm, representation independence for
    substitutions and typing contexts, heterogeneous equality between
    renamings and substitutions, and absence of postulated axioms.
  \item
    three case studies using our approach to prove subject reduction
    for System~F with subtyping, a dependently-typed lambda calculus, and
    pattern matching.
  \end{enumerate}

@@

  \section{Subject Reduction}
  \label{sec:subjectreduction}

  We define a standard small-step operational semantics for System F
  with full reduction:
  \AppCode\FReduction

  Proving subject reduction is now straightforward:
  \AppCode\FSubjectReduction
  \AppCode\FSubjectReductionProofInteresting
  We only listed the cases that do not follow directly from the
  induction hypothesis:
  \begin{itemize}
  \item The beta-reduction rule for type-application, \ACode{\ACon{β-Λ}},
    is a direct consequence of substitution preserves typing.
    \ACode{⊢e₂} is the typing for \ACode{e₂}.
    \ACode{⊢⦅ ⊢e₂ ⦆} is the substitution-typing for the singleton substitution 
    \ACode{\AField{⦅ e₂ ⦆}}.
    \ACode{\AField{⊢e₁ ⊢⋯ ⊢⦅ ⊢e₂ ⦆}} states that applying the singleton substitution
    to \ACode{e₁} preserves typing.
  \item The beta-reduction rule for value-application is almost the same,
    but requires us to addionaly apply an equality to the type index.
  \end{itemize}
